Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-13 BN_discretizer_paper.tex
cleaned up intro, removed tim comments, fixed notation section, replaced double verts with single

There are two common approaches to extending Bayesian networks to continuous variables. The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters. One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}. For other particle shapes \citep{Ihler_2009} and non-linear functions~\todo{[cite]}, these redesigned algorithms are computationally expensive and do not perform well.
\tim{Not sure what `particle shapes' are. Do you mean other parametric distributions?}

@@

This paper describes a new Bayesian discretization method for continuous variables in Bayesian networks, extending prior work on single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.
The proposed method optimizes the discretization policy relative to the network and takes parents, children, and spouse variables into account.
The optimal single-variable discretization method is derived in Section \ref{sec:single_var}, and a second heuristic method is proposed which reduces the exponential runtime to polynomial.
Section \ref{sec:multi_var} covers Bayesian networks with multiple continuous variables and discretization while simultaneously learning network structure.
The paper concludes with a comparison against existing minimum-description length \citep{Friedman_1996} and MODL \citet{Boulle_2006} methods on real-world datasets in Section \ref{sec:experiments}.

@@

\section{Related Work}
\label{sec:related_work}
This section reviews the minimum description length (MDL) principle discretization \citep{Friedman_1996} and MODL discretization \citep{Boulle_2006}.
The former is the predominant discretization method for continuous variables in Bayesian networks, and is compared to our proposed method in Section \ref{sec:experiments}.
The latter is a discretization method for one continuous variable against a discrete target class, which our proposed method generalizes to Bayesian networks.
The asymptotical equivalence between MDL and MODL was examined in \citep{VL_2000}.

@@

\begin{small}
\begin{equation}
\begin{aligned}
\frac{1}{2} \log(N) \left\lbrace  \lVert \Pi_{X_i} \rVert (\lVert X^*_i \rVert - 1) +
 {\sum_{j,X_i \in \Pi_{X_i}}} \lVert \Pi_{X^*_j} \rVert (\lVert X_j \rVert - 1) \right\rbrace + \log(\lVert {X^*_i} \rVert) \\
 + (N_i-1) H \left( {{\|X^*_i\| - 1}\over{N_i -1}}  \right) -N \cdot \left[ I(X^*_i,\Pi_{X_i}) + {\sum_{j,X_i \in \Pi_{X_i}}} I(X_j, \Pi_{X^*_j}) \right],
\end{aligned}
\end{equation}
\end{small}

\noindent
% $I(\boldsymbol{A},\boldsymbol{B}) =  \sum_{\boldsymbol{a},\boldsymbol{b}} \hat{P}_D (\boldsymbol{a},\boldsymbol{b}) \log {{\hat{P}_D (\boldsymbol{a},\boldsymbol{b})}\over{\hat{P}_D (\boldsymbol{a}),\hat{P}_D (\boldsymbol{b})}}$
where $I(\boldsymbol{A},\boldsymbol{B})$ is the mutual information between the discrete variables $\boldsymbol{A}$ and $\boldsymbol{B}$, $\|\boldsymbol{X}\|$ is the cardinality of discrete variable $\boldsymbol{X}$, $X^*_i$ is the discretized version of continuous variable $X_i$, and $H(p) = -p \log(p) - (1-p) \log(1-p)$.
The optimal discretization policy can be found with dynamical programming.
For Bayesian networks with multiple continuous variables, the algorithm is applied to one variable at a time iteratively over all continuous variables.
While iterating, only one variable is treated as continuous and other continuous variables are discretized based on an initial discretization policy or the discretization result from a previous iteration.

@@

\subsection{MODL Discretization}
\tim{Does MODL stand for something?}
MODL discretization \citep{Boulle_2006} is a Bayesian method for discretizing a continuous feature according to a class variable. Bayesian methods maximize the probability of the model given the data, $P(\textit{Model} \mid \textit{Data})$. An application of Bayes' rule and constant $P(\textit{Data})$ shows that this is equivalent to maximizing:
\begin{align}
P(\textit{Model}) \cdot P(\textit{Data} \mid \textit{Model})\text{,}
\end{align}

\noindent
where $P(\textit{Model})$ is a prior over models and $P(\textit{Data} \mid \textit{Model})$ is the data likelihood.
The data likelihood usually increases with the number of discretization intervals, as more intervals allow for more detailed representations.
The prior usually decreases as the number of discretization intervals increases to favor simpler models and penalize overfitting.
Thus, maximizing $P(\textit{Model}) \cdot P(\textit{Data} \mid \textit{Model})$ requires a trade-off when determining the number of discretization intervals.

The MODL method uses dynamic programming to find the optimal discretization policy, and has a $\bigo\paren{N^3 + \|\text{Class}\| \cdot N^2}$ runtime, where $N$ is the of number of data instances. \cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$\tim{Did I state this correctly?}. If the prior is chosen to reflect uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{\|\text{Class}\| \cdot N^2}$.\tim{Is a sacrifice made? It is sub-optimal, right?}

@@

A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variables $X = \brock{X_1, X_2, \cdots, X_n}$, and whose edges $E$ represent probabilistic dependencies among nodes. The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pi{X_i}$ in $G$. The conditional relations are parameterized by the elements of $\Theta$, which take the form $\theta_{x_i \mid \Pi_{x_i}} = P(x_i \mid \Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$. The joint distribution over $X$ factors according to the Markov property:
\begin{equation}
P_B (X_1 , \cdots, X_n) = \prod_{i=1}^{n} P_B (X_i \mid \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i \mid \Pi_{x_i}}\text{.}
\end{equation}

Samples can be drawn from the joint distribution represented by a Bayesian network by forward sampling.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
Furthermore, the parent\rightarrow child sampling order can be reversed if the marginal probability of child variable $P(X_i)$ is known.\tim{Do we use this later?}
By Bayes' rule,

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based structure learning, score-based structure learning, and Bayesian model averaging \citep[see][chap.~18]{PGM_2009}. This work uses the K2 structure learning algorithm \citep{K2}, one of the most successful score-based structure learning methods.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures are scored according to their likelihood, $\prod_i f(X_i, \Pi_{X_i})$, where

\begin{align}
f(X_i, \Pi_{X_i}) = \prod_{j=1}^{\|\Pi_{X_i}\|} {{(\|X_i\|-1)!}\over{(N_{ij} + \|X_i\|-1)!}} \prod_{k=1}^{\|X_i\|} \alpha_{ijk}!\text{.}
\end{align}

\noindent
$\|X_i\|$ is the cardinality of $X_i$, $\|\Pi_{X_i}\|$ is number of possible instantiations of the parent variables of $X_i$, $\alpha_{ijk}$ is the number of instances of the $k$th value of $X_i$ and the $j$th parental instantiation, and $N_{ij} = \sum_{k=1}^{|X_i|} \alpha_{ijk}$.\tim{Mykel uses $\alpha_{ij0}$}
The optimal structure is the most probably structure given the data.
The space of acyclic graphcs is superexponential in the number of nodes \citep{PGM_2009} so it is common to rely on heuristic search strategies.

The K2 algorithm assumes a topological ordering of variables and greedily adds parents to nodes to maximally increase the likelihood score.
A fixed ordering ensures acyclicity but does not guarantee a globally optimal network structure.
K2 is typically run multiple times with random topological orderings and the network with the highest likelihood is retained.

@@

\noindent
The policy discretizes the continuous variable $C$ into $k$ intervals. Furthermore, we restrict the discretization edges $t_\cdot$ to midpoints between two sorted instances of $C$. An integer representation of $M_C$ can be obtained by sorting the samples of $C$ in an increasing order, $\curly{c_1,c_2,\ldots,c_N}$, and if $t_i = (c_{s_{i}} + c_{s_{i}+1})/2$ for $i=1,2,\ldots,k$, then $M_C = \anglebrackets{t_1,t_2, \ldots, t_{k-1}} \equiv \brock{n_1,n_2,..n_k}$, where $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ for all $i = 2,\ldots,k$.

@@

This section covers discretization of a single continuous variable in a Bayesian network where all other variables are discrete.
Before we formulate the objective function, we introduce some notations that will make the following calculation easier.

\subsection{Notations}

\tim{We should move the notation section to preliminaries}

\tim{Earlier the continuous variable was denoted $C$. Now it is $X$. Slightly confusing with $c$ also meaning child.}

Consider continuous variable $X$.
It has $n_p$ parents, $\Pi_X = \{ P_1, P_2, P_3,\ldots,P_{n_p}\}$ and $n_c$ children, $\{ C_1, C_2,\ldots,C_{n_c}\}$.
Each child $C_i$ of $X$ may possess other parents, our spouses of $X$, whose set is denoted $\boldsymbol{S_i}$.

Let $D$ be a dataset of $N$ joint instances from which discretization will be learned, sorted in ascending order according to $X$.
Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$.
Thus $D_X = \{ x_1,x_2,x_3,\ldots,x_N \}$.
In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_N$.
It follows that a discretization policy $M$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.
The optimal discretization policy maximizes $P(M)\cdot P(D\mid M)$.

@@

For the following part, we evaluate $P(D_{\Pi_x} \mid M)$  and $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$ based on discretization policy $M = [n_1,n_2,\ldots,n_k]$ and dataset $D$.
\subsubsection{Evaluate $P(D_{\Pi_x} \mid M)$}
Assume that $J_P = \prod_{i=1}^{n_p} \| P_i \|$. Then we have:
\begin{equation}
P(D_{\Pi_X} \mid M) = \prod_{i=1}^k  {{1}\over{{n_i + J_P - 1}\choose{J_P -1}}}
{{1}\over{ {{{n}_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}  }},
\end{equation}
where $n^{(p)}_{i,j}$ is the number of instances in $i$th discretized interval with $j$th value of $\Pi_X$. Note that $n_i = \sum_{j=1}^{\| \Pi_x \|} n^{(p)}_{i,j}$. The two factors on RHS comes from the second prior: all distributions of values of $\Pi_X$ in a given interval are equiprobable. According to the forth prior, the distribution in each interval is independent, so we multiply all the two factors together.
\subsubsection{Evaluate $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$}
Assume for each pair of $(C_j, \boldsymbol{S_j})$, we have $\|C_j \| = J_j$ and $\| \boldsymbol{S_j} \| = L_j = \prod_{v \in \boldsymbol{S_j}} \| v \|$. Therefore,
\begin{equation}
P(D_{C_j}  \mid M, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,l}$ is the number of instances in $i$th interval with $m$th value of $C_j$ and $l$th value of $\boldsymbol{S_j}$, $n_{i,l} = \sum_{m=1}^{J_j} n^{(j)}_{i,m,l}$, and $n_i = \sum_{l=1}^{L_j} n_{i,l}$. The two factors on RHS comes from the third prior: all distribution of values of $C_j$ in a given interval and with a given value of $S_j$ are equiprobable. According to the forth prior, these distributions are independent from each other, therefore we multiply all factors. If $S_j = \emptyset$, then Equation 13. is equivalent to
\begin{equation}
P(D_{C_j}  \mid M) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,\emptyset}$ is the number of instances in $i$th interval with $m$th value of $C_j$, and $\sum_{m=1}^{J_j} n^{(j)}_{i,m,\emptyset} = n_i$. \\

@@

The evaluation of function $h(u,v)$ for all $u \leq v$ is summerized in Algorithm 4 in Appendix. The calculation can be done in $O(n_c  {v'_{max}} \cdot N^2 + {v'_{max}}^{n_p} \cdot N^2)$, where $n_c$ and $n_p$ are the numbers of child and parent variables, respectively, and $v'_{max}$ is the largest cardinality of variables that directly connects to $X$. Notice that due to repeated values in $D_X$, for some pairs of $u$ and $v$, the value of $h(u,v)$ might depend on the sorting method of $D_X$ and $D$. However, this does not influence the optimization result, since these pairs of $u$ and $v$ will not form valid intervals.

Now we are able to solve the optimization problem with the objective function in Equation 14. The dynamical programming procedure is shown in Algorithm 1. We have three inputs: $D_X$, the data of continuous variable $X$ in the ascending order, $D$, the data of other variables and sorted according to $D_X$, and $G$, the network structure. The runtime of Algorithm 1 is also $O(n_c  {v'_{max}} \cdot N^2 + {v'_{max}}^{n_p} \cdot N^2)$, since the dynamical programming procedure does not increase the time complexity beyond the calculation of $h(u,v)$. There are other methods that lead to suboptimal results but runs faster than dynamical programming method. Please refer to \citep{Boulle_2006}.

@@

\subsection{Approximation}
\label{approx}
Notice that Algorithm1 leads to the non-polynomial runtime $O({v'_{max}}^{n_p} \cdot N^2)$ in the discretization process. Therefore, we have an empirical approximation to the objective function that can reduce runtime and still preserve the quality of discretization. The approximation is as follows: for the dominator of the last factor in Equation 11, we have
\begin{equation}
{{{{n}_i}!}\over{ {{n^{(p)}}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}} \approx \prod_{r=1}^{n_p} { {{{n}_i}!}\over{ {n^{(p_r)}_{i,1} !} {n^{(p_r)}_{i,2} !} \cdots {n^{(p_r)}_{i,J_{p_r}} !}}},
\end{equation}
where $J_{p_r} = \| P_r\|$ and ${n^{(p_r)}_{i,j} !}$ is the number of instances in $i$th interval with $j$th value of $P_r$. Applying this approximation, the approximated objective function is
\begin{equation}
\begin{aligned}
& \sum_{i=1}^{k-1} - \log(1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}) +  \sum_{i=1}^{k} L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}} +\\
&  \sum_{j=1}^{n_c} \sum_{i=1}^{k}  \sum_{l=1}^{L_j} \left[  \log{{n_{i,l} + J_j - 1}\choose{J_j-1}} + \log \left( { {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}} } \right) \right] + \\
& \sum_{i=1}^k \left[  \log {{n_{i} + J_P - 1}\choose{J_P-1}} + \sum_{q=1}^{n_p} \log \left( { {{{n}_i}!}\over{ {n^{(p_q)}_{i,1} !} {n^{(p_q)}_{i,2} !} \cdots {n^{(p_q)}_{i,J_{p_q}} !}}} \right) \right].
\end{aligned}
\end{equation}

As we will see in Section \ref{sec:experiments}, the approximated algorithm is more sensitive to the distribution of other variables' values and usually leads to slightly more discretization edges. With the approximation, we can reduce runtime of Algorithm 1 from $O(n_c  {v'_{max}} \cdot N^2 + {v'_{max}}^{n_p} \cdot N^2)$ to $O( {v'_{max}} (n_c + n_p) \cdot N^2)$. The rigorous proof of the approximation have not been proposed.

@@

\begin{algorithm}
\caption{ Calculation of function $h(u,v)$ for all $u \leq v$}\label{euclid}
\begin{algorithmic}[1]
\State Initialize $H$ as an $N \times N$ matrix that all elements are 0.
\State $count_p$ is an $N \times N \times \|\Pi_X\|$ matrix such that $count_p [u,v,w]$ is the number of instances from $x_u$ to $x_v$ with $w$th value of $\Pi_X$. This matrix can be calculated in $O(\|\Pi_X\| \cdot N^2)$
\For {$u = 1$ to $N$}
\For {$v = u$ to $N$}
\State $H(u,v) \leftarrow H(u,v) + \log((v-u + J_p)!) - \log((J_p -1)!)$
\For {$w = 1$ to $\|\Pi_X\|$}
\State $H(u,v) \leftarrow H(u,v) - \log( count_p(u,v,w)!)$
\EndFor
\EndFor
\EndFor

>>>>

There are two common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions \citep{Ihler_2009} and non-linear functions~\todo{[cite]}, these redesigned algorithms are computationally expensive and do not perform well.

@@

This paper describes a new Bayesian discretization method for continuous variables in Bayesian networks, extending prior work on single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.
The proposed method optimizes the discretization policy relative to the network and takes parents, children, and spouse variables into account.
The optimal single-variable discretization method is derived in Section \ref{sec:single_var}, and a second heuristic method is proposed which reduces the runtime from exponential to polynomial.
Section \ref{sec:multi_var} covers Bayesian networks with multiple continuous variables and discretization while simultaneously learning network structure.
The paper concludes with a comparison against the existing minimum-description length \citep{Friedman_1996} method on real-world datasets in Section \ref{sec:experiments}.

@@

\section{Related Work}
\label{sec:related_work}
This section reviews the minimum description length (MDL) principle discretization \citep{Friedman_1996} and MODL discretization \citep{Boulle_2006} methods.
The former is the predominant discretization method for continuous variables in Bayesian networks, and is compared to our proposed method in Section \ref{sec:experiments}.
The latter is a discretization method for one continuous variable against a discrete target class, which our proposed method generalizes to Bayesian networks.
The asymptotical equivalence between MDL and MODL was examined in \citep{VL_2000}.

@@

% \begin{small}
\begin{equation}
\begin{aligned}
\frac{1}{2} \log(N) \left\lbrace  \lvert \Pi_{X_i} \rvert (\lvert X^*_i \rvert - 1) +
 {\sum_{j,X_i \in \Pi_{X_i}}} \lvert \Pi_{X^*_j} \rvert (\lvert X_j \rvert - 1) \right\rbrace + \log(\lvert {X^*_i} \rvert) \\
 + (N_i-1) H \left( {{|X^*_i| - 1}\over{N_i -1}}  \right) -N \cdot \left[ I(X^*_i,\Pi_{X_i}) + {\sum_{j,X_i \in \Pi_{X_i}}} I(X_j, \Pi_{X^*_j}) \right],
\end{aligned}
\end{equation}
% \end{small}

\noindent
% $I(\boldsymbol{A},\boldsymbol{B}) =  \sum_{\boldsymbol{a},\boldsymbol{b}} \hat{P}_D (\boldsymbol{a},\boldsymbol{b}) \log {{\hat{P}_D (\boldsymbol{a},\boldsymbol{b})}\over{\hat{P}_D (\boldsymbol{a}),\hat{P}_D (\boldsymbol{b})}}$
where $I(\boldsymbol{A},\boldsymbol{B})$ is the mutual information between the discrete variables $\boldsymbol{A}$ and $\boldsymbol{B}$, ${H(p) = -p \log(p) - (1-p) \log(1-p)}$, $|\boldsymbol{X}|$ is the cardinality of discrete variable $\boldsymbol{X}$, and $X^*_i$ is the discretized version of the $i$th sample of continuous variable $X$.
The optimal discretization policy can be found with dynamical programming.
For Bayesian networks with multiple continuous variables, the algorithm is applied to one variable at a time iteratively over all continuous variables.
While iterating, only one variable is treated as continuous and other continuous variables are discretized based on an initial discretization policy or the discretization result from a previous iteration.

@@

\subsection{MODL Discretization}
MODL discretization \citep{Boulle_2006} is a Bayesian method for discretizing a continuous feature according to a class variable. Bayesian methods maximize the probability of the model given the data, $P(\textit{Model} \mid \textit{Data})$. An application of Bayes' rule and constant $P(\textit{Data})$ shows that this is equivalent to maximizing:
\begin{align}
P(\textit{Model}) \cdot P(\textit{Data} \mid \textit{Model})\text{,}
\end{align}

\noindent
where $P(\textit{Model})$ is a prior over models and ${P(\textit{Data} \mid \textit{Model})}$ is the data likelihood.
The data likelihood usually increases with the number of discretization intervals, as more intervals allow for richer representations.
The prior decreases as the number of discretization intervals increases to favor simpler models and prevent overfitting.
Maximizing ${P(\textit{Model}) \cdot P(\textit{Data} \mid \textit{Model})}$ requires a trade-off when determining the number of discretization intervals.

The MODL method uses dynamic programming to find the optimal discretization policy, and has a $\bigo\paren{N^3 + |\text{Class}| \cdot N^2}$ runtime, where $N$ is the of number of data instances.
\cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$.
Under a uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{|\text{Class}| \cdot N^2}$ without sacrificing optimality.

@@

A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variables $X = \brock{X_1, X_2, \cdots, X_n}$, and whose edges $E$ represent probabilistic dependencies among nodes. The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pi_{X_i}$ in $G$. The conditional relations are parameterized by the elements of $\Theta$, which take the form $\theta_{x_i \mid \Pi_{x_i}} = P(x_i \mid \Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$. The joint distribution over $X$ factors according to the Markov property:
\begin{equation}
P_B (X_1 , \cdots, X_n) = \prod_{i=1}^{n} P_B (X_i \mid \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i \mid \Pi_{x_i}}\text{.}
\end{equation}

Forward sampling is used to draw samples from the joint distribution represented by a Bayesian network.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
Furthermore, the parent\rightarrow child sampling order can be reversed if the marginal probability of the child variable is known.
By Bayes' rule,

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based, score-based, and Bayesian model averaging \citep[see][chap.~18]{PGM_2009}. This work uses the K2 structure learning algorithm \citep{K2}, one of the most successful score-based structure learning methods.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures according to their likelihood, $\prod_i f(X_i, \Pi_{X_i})$, where

\begin{align}
f(X_i, \Pi_{X_i}) = \prod_{j=1}^{|\Pi_{X_i}|} {{(|X_i|-1)!}\over{(N_{ij} + |X_i|-1)!}} \prod_{k=1}^{|X_i|} \alpha_{ijk}!\text{,}
\end{align}

\noindent
where $|X_i|$ is the cardinality of $X_i$, $|\Pi_{X_i}|$ is number of possible instantiations of the parent variables of $X_i$, $\alpha_{ijk}$ is the number of instances of the $k$th value of $X_i$ with the $j$th parental instantiation, and $N_{ij} = \sum_{k=1}^{|X_i|} \alpha_{ijk}$.
The optimal structure is the most probably structure given the data.
The space of acyclic graphcs is superexponential in the number of nodes so it is common to rely on heuristic search strategies \citep{PGM_2009}.

The K2 algorithm assumes a topological ordering of variables and greedily adds parents to nodes to maximally increase the likelihood score.
A fixed ordering ensures acyclicity but does not guarantee a globally optimal network structure.
K2 is typically run multiple times with random variable orderings and the network with the highest likelihood is retained.

@@

\noindent
The policy discretizes the continuous variable $C$ into $k$ intervals.
The discretization edges $t_\cdot$ are restricted to the midpoints between two sorted instances of $C$.
An integer representation of $M_C$ can be obtained by sorting the samples of $C$ in an increasing order, $\curly{c_1,c_2,\ldots,c_N}$, and if $t_i = (c_{s_{i}} + c_{s_{i}+1})/2$ for $i=1,2,\ldots,k$, then $M_C = \anglebrackets{t_1,t_2, \ldots, t_{k-1}} \equiv \brock{n_1,n_2,..n_k}$, where $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$.

@@

This section covers discretization of a single continuous variable $X$ in a Bayesian network where all other variables are discrete.
Let $n_p$ and $n_c$ respectively refer to the number of parents and number of children of $X$.
Each child $C_i$ of $X$ may possess other parents, referred to as spouses of $X$.
The set of spouses of $X$ is denoted $\boldsymbol{S_i}$.

Let $D$ be a dataset of $N$ samples from which the discretization will be learned, sorted in ascending order according to $X$.
Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$, thus $D_X = \{ x_1,x_2,x_3,\ldots,x_N \}$.
In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_N$.
It follows that a discretization policy $M$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.
An optimal discretization policy maximizes $P(M)\cdot P(D\mid M)$ for some prior $P(M)$ and likelihood $P(D\mid M)$.

@@

For the following part, we evaluate $P(D_{\Pi_x} \mid M)$  and $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$ based on discretization policy $M = [n_1,n_2,\ldots,n_k]$ and dataset $D$.
\subsubsection{Evaluate $P(D_{\Pi_x} \mid M)$}
Assume that $J_P = \prod_{i=1}^{n_p} | P_i |$. Then we have:
\begin{equation}
P(D_{\Pi_X} \mid M) = \prod_{i=1}^k  {{1}\over{{n_i + J_P - 1}\choose{J_P -1}}}
{{1}\over{ {{{n}_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}  }},
\end{equation}
where $n^{(p)}_{i,j}$ is the number of instances in $i$th discretized interval with $j$th value of $\Pi_X$. Note that $n_i = \sum_{j=1}^{| \Pi_x |} n^{(p)}_{i,j}$. The two factors on RHS comes from the second prior: all distributions of values of $\Pi_X$ in a given interval are equiprobable. According to the forth prior, the distribution in each interval is independent, so we multiply all the two factors together.
\subsubsection{Evaluate $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$}
Assume for each pair of $(C_j, \boldsymbol{S_j})$, we have $|C_j | = J_j$ and $| \boldsymbol{S_j} | = L_j = \prod_{v \in \boldsymbol{S_j}} | v |$. Therefore,
\begin{equation}
P(D_{C_j}  \mid M, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,l}$ is the number of instances in $i$th interval with $m$th value of $C_j$ and $l$th value of $\boldsymbol{S_j}$, $n_{i,l} = \sum_{m=1}^{J_j} n^{(j)}_{i,m,l}$, and $n_i = \sum_{l=1}^{L_j} n_{i,l}$. The two factors on RHS comes from the third prior: all distribution of values of $C_j$ in a given interval and with a given value of $S_j$ are equiprobable. According to the forth prior, these distributions are independent from each other, therefore we multiply all factors. If $S_j = \emptyset$, then Equation 13. is equivalent to
\begin{equation}
P(D_{C_j}  \mid M) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,\emptyset}$ is the number of instances in $i$th interval with $m$th value of $C_j$, and $\sum_{m=1}^{J_j} n^{(j)}_{i,m,\emptyset} = n_i$. \\

@@

The evaluation of function $h(u,v)$ for all $u \leq v$ is summerized in Algorithm 4 in Appendix. The calculation can be done in $\bigo\paren{n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2}$, where $n_c$ and $n_p$ are the numbers of child and parent variables, respectively, and $v'_\text{max}$ is the largest cardinality of variables that directly connects to $X$. Notice that due to repeated values in $D_X$, for some pairs of $u$ and $v$, the value of $h(u,v)$ might depend on the sorting method of $D_X$ and $D$. However, this does not influence the optimization result, since these pairs of $u$ and $v$ will not form valid intervals.

Now we are able to solve the optimization problem with the objective function in Equation 14. The dynamical programming procedure is shown in Algorithm 1. We have three inputs: $D_X$, the data of continuous variable $X$ in the ascending order, $D$, the data of other variables and sorted according to $D_X$, and $G$, the network structure. The runtime of Algorithm 1 is also $\bigo\paren{n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2}$, since the dynamical programming procedure does not increase the time complexity beyond the calculation of $h(u,v)$. There are other methods that lead to suboptimal results but runs faster than dynamical programming method. Please refer to \citep{Boulle_2006}.

@@

\subsection{Approximation}
\label{approx}
Notice that Algorithm1 leads to the non-polynomial runtime $\bigo\paren{{v'_\text{max}}^{n_p} \cdot N^2}$ in the discretization process. Therefore, we have an empirical approximation to the objective function that can reduce runtime and still preserve the quality of discretization. The approximation is as follows: for the dominator of the last factor in Equation 11, we have
\begin{equation}
{{{{n}_i}!}\over{ {{n^{(p)}}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}} \approx \prod_{r=1}^{n_p} { {{{n}_i}!}\over{ {n^{(p_r)}_{i,1} !} {n^{(p_r)}_{i,2} !} \cdots {n^{(p_r)}_{i,J_{p_r}} !}}},
\end{equation}
where $J_{p_r} = | P_r|$ and ${n^{(p_r)}_{i,j} !}$ is the number of instances in $i$th interval with $j$th value of $P_r$. Applying this approximation, the approximated objective function is
\begin{equation}
\begin{aligned}
& \sum_{i=1}^{k-1} - \log(1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}) +  \sum_{i=1}^{k} L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}} +\\
&  \sum_{j=1}^{n_c} \sum_{i=1}^{k}  \sum_{l=1}^{L_j} \left[  \log{{n_{i,l} + J_j - 1}\choose{J_j-1}} + \log \left( { {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}} } \right) \right] + \\
& \sum_{i=1}^k \left[  \log {{n_{i} + J_P - 1}\choose{J_P-1}} + \sum_{q=1}^{n_p} \log \left( { {{{n}_i}!}\over{ {n^{(p_q)}_{i,1} !} {n^{(p_q)}_{i,2} !} \cdots {n^{(p_q)}_{i,J_{p_q}} !}}} \right) \right].
\end{aligned}
\end{equation}

As we will see in Section \ref{sec:experiments}, the approximated algorithm is more sensitive to the distribution of other variables' values and usually leads to slightly more discretization edges. With the approximation, we can reduce runtime of Algorithm 1 from $\bigo\paren{n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2}$ to $\bigo\paren{v'_\text{max} (n_c + n_p) \cdot N^2}$. The rigorous proof of the approximation have not been proposed.

@@

\begin{algorithm}
\caption{ Calculation of function $h(u,v)$ for all $u \leq v$}\label{euclid}
\begin{algorithmic}[1]
\State Initialize $H$ as an $N \times N$ matrix that all elements are 0.
\State $count_p$ is an $N \times N \times |\Pi_X|$ matrix such that $count_p [u,v,w]$ is the number of instances from $x_u$ to $x_v$ with $w$th value of $\Pi_X$. This matrix can be calculated in $\bigo\paren{|\Pi_X| \cdot N^2}$
\For {$u = 1$ to $N$}
\For {$v = u$ to $N$}
\State $H(u,v) \leftarrow H(u,v) + \log((v-u + J_p)!) - \log((J_p -1)!)$
\For {$w = 1$ to $|\Pi_X|$}
\State $H(u,v) \leftarrow H(u,v) - \log( count_p(u,v,w)!)$
\EndFor
\EndFor
\EndFor
